---
title: "pilot_3_data_analysis"
output: html_document
date: "2026-04-20"
---

```{r}
# load packages

library(tidyverse)
library(ggbeeswarm)
library(brms)
library(tidyboot)
library(tidyjson)
library(tidybayes)
library(patchwork)
library(GGally)
library(cowplot)
library(BayesFactor)
library(aida)   # custom helpers: https://github.com/michael-franke/aida-package
library(faintr) # custom helpers: https://michael-franke.github.io/faintr/index.html
library(cspplot)
library(knitr)
library(ggplot2)
library(dplyr)
library(gtools)
library(Hmisc)
library(purrr)
library(purrr)
library(bayestestR)


##################################################

# these options help Stan run faster
options(mc.cores = parallel::detectCores(),
        brms.backend = "cmdstanr")

# use the CSP-theme for plotting
theme_set(theme_csp())

# global color scheme from CSP
project_colors = cspplot::list_colors() |> pull(hex)

# setting theme colors globally
scale_colour_discrete <- function(...) {
  scale_colour_manual(..., values = project_colors)
}

scale_fill_discrete <- function(...) {
  scale_fill_manual(..., values = project_colors)
}

##################################################

# open data from pilot3 
data <- read.csv("../pilot_3_data/A1-Indicom-pilot-v3-results-collapsed.csv", sep=";") 

# filter out control -> leave only main trials
df_main <- data[data$condition !='control',]

head(df_main)
```

## 

```{r}

#trim stray whitespace
df_main$contextType       <- trimws(df_main$contextType)
df_main$informationStatus <- trimws(df_main$informationStatus)

#ensure correct types
df_main$rating            <- as.numeric(df_main$rating)
df_main$informationStatus <- factor(df_main$informationStatus, levels = c(
  "at-issue-explicit", "at-issue-implicit", "no-longer-at-issue", "never-at-issue"
))
df_main$contextType <- factor(df_main$contextType, levels = c("positive", "negative"))


#get means, median and mode
mode_val <- function(x) {
  x <- x[!is.na(x)]
  as.numeric(names(sort(table(x), decreasing = TRUE))[1])
}
 
summary_table <- df_main %>%
  group_by(informationStatus, contextType) %>%
  summarise(
    n      = n(),
    mean   = round(mean(rating, na.rm = TRUE), 2),
    median = median(rating, na.rm = TRUE),
    mode   = mode_val(rating),
    sd     = round(sd(rating, na.rm = TRUE), 2),
    .groups = "drop"
  ) %>%
  arrange(informationStatus, contextType)
 
print(summary_table, n = Inf)
write.csv(summary_table, "summary_table.csv", row.names = FALSE)
message("Summary table saved to summary_table.csv")





```



```{r }

df_main$informationStatus <- recode(df_main$informationStatus,
  "at-issue-explicit"  = "at-issue\nexplicit",
  "at-issue-implicit"  = "at-issue\nimplicit",
  "no-longer-at-issue" = "no-longer\nat-issue",
  "never-at-issue"     = "never\nat-issue"
)


```
```{r }
# ── Plot 2: x = informationStatus, grouped by contextType ───────────────────
p2 <- ggplot(df_main, aes(x = informationStatus, y = rating, fill = contextType)) +
  stat_summary(
    fun = mean,
    geom = "bar",
    position = position_dodge(0.9)
  ) +
  stat_summary(
    fun.data = mean_cl_boot,
    geom = "errorbar",
    width = 0.2,
    position = position_dodge(0.9)
  ) +
  labs(
    x     = "Information Status",
    y     = "Mean Rating",
    fill  = "Context Type",
    title = "Mean Ratings by Information Status and Context Type\n(95% Bootstrapped Confidence Intervals)"
  ) +
  theme_csp()
 
print(p2)
ggsave("plot2_across_informationStatus.png", plot = p2, width = 10, height = 6, dpi = 150)
```






```{r }


# ── Participant mean ratings (overall) ───────────────────────────────────────
pp_summary <- df_main %>%
  group_by(submission_id) %>%
  summarise(
    mean_rating = mean(rating, na.rm = TRUE),
    lower_ci    = smean.cl.boot(rating)[2],
    upper_ci    = smean.cl.boot(rating)[3],
    n           = n(),
    .groups = "drop"
  )
 
ggplot(pp_summary, aes(x = reorder(submission_id, mean_rating), y = mean_rating)) +
  geom_point() +
  geom_errorbar(aes(ymin = lower_ci, ymax = upper_ci), width = 0.2) +
  coord_flip() +
  theme_csp() +
  labs(
    x     = "Participant",
    y     = "Mean Rating",
    title = "Participant Mean Ratings"
  )
 
ggsave("plot_pp_overall.png", width = 8, height = 6, dpi = 150)

```

```{r }
# ── Participant mean ratings by contextType ───────────────────────────────────
pp_context_summary <- df_main %>%
  group_by(submission_id, contextType) %>%
  summarise(
    mean_rating = mean(rating, na.rm = TRUE),
    lower_ci    = smean.cl.boot(rating)[2],
    upper_ci    = smean.cl.boot(rating)[3],
    .groups = "drop"
  )
 
ggplot(pp_context_summary, aes(x = contextType, y = mean_rating)) +
  geom_point() +
  geom_errorbar(aes(ymin = lower_ci, ymax = upper_ci), width = 0.2) +
  facet_wrap(~ submission_id) +
  theme_minimal() +
  labs(
    x     = "Context Type",
    y     = "Mean Rating",
    title = "Participant Mean Ratings by Context Type"
  )
 
ggsave("plot_pp_by_context.png", width = 12, height = 8, dpi = 150)

```

```{r}
df_main$informationStatus <- recode(df_main$informationStatus,
  "at-issue\nexplicit"  = "at_issue_explicit",
  "at-issue\nimplicit"  = "at_issue_implicit",
  "no-longer\nat-issue" = "no_longer_at_issue",
  "never\nat-issue"     = "never_at_issue"
)
```



```{r, message=FALSE, warning=FALSE}


 
df_main$informationStatus <- recode(df_main$informationStatus,
  "at-issue\nexplicit"  = "at_issue_explicit",
  "at-issue\nimplicit"  = "at_issue_implicit",
  "no-longer\nat-issue" = "no_longer_at_issue",
  "never\nat-issue"     = "never_at_issue"
)


# ── Model ────────────────────────────────────────────────────────────────────
# informationStatus is the main predictor
# contextType included as a control variable
# random intercepts for participants and items
 
model <- brm(
  #rating ~ informationStatus + contextType + (1 | submission_id) + (1 | itemID),
    rating ~ informationStatus * contextType + (1 + informationStatus * contextType | submission_id) + (1 | itemID),
  data   = df_main,
  family = cumulative(link = "logit"),
  chains = 4,
  cores  = 4,
  iter   = 4000
)
 
summary(model)
saveRDS(model, "pilot_model.rds")
 
```
```{r}

draws <- df_main  %>% 
  select(informationStatus, contextType) %>% 
  unique() %>% 
  tidybayes::add_linpred_draws( #use add_epred_draws() instead? -> expected ratings instead of log odds
    model,
    ndraws = 2000,#number of draws
    re_formula = NA
  )

```


```{r}
deltas <- draws %>%
  ungroup() %>%
  select(.draw, informationStatus, contextType, .linpred) %>%
  pivot_wider(
    names_from = contextType,
    values_from = .linpred
  ) %>%
  mutate(
    delta = positive - negative #compute the difference between positive and negative per conditions
  )
```

```{r}
myVec_at_issue_explicit <- deltas %>%
  filter(informationStatus == "at_issue_explicit") %>%
  pull(delta)

myVec_at_issue_implicit <- deltas %>%
  filter(informationStatus == "at_issue_implicit") %>%
  pull(delta)

myVec_no_longer_at_issue <- deltas %>%
  filter(informationStatus == "no_longer_at_issue") %>%
  pull(delta)

myVec_never_at_issue <- deltas %>%
  filter(informationStatus == "never_at_issue") %>%
  pull(delta)
```


```{r}
hist(myVec_at_issue_explicit)
hist(myVec_at_issue_implicit)
hist(myVec_no_longer_at_issue)
hist(myVec_never_at_issue)

aida::summarize_sample_vector(myVec_at_issue_explicit)
aida::summarize_sample_vector(myVec_at_issue_implicit)
aida::summarize_sample_vector(myVec_no_longer_at_issue)
aida::summarize_sample_vector(myVec_never_at_issue)


```
```{r}
delta_comparisons <- deltas %>%
  select(.draw, informationStatus, delta) %>%
  pivot_wider(
    names_from = informationStatus,
    values_from = delta
  ) %>%
  mutate(
    explicit_vs_nolonger =
      at_issue_explicit - no_longer_at_issue,

    explicit_vs_implicit =
      at_issue_explicit - at_issue_implicit,

    nolonger_vs_never =
      no_longer_at_issue - never_at_issue,
    
    explicit_vs_never = 
      at_issue_explicit - never_at_issue
  )
```

```{r}
myVec_explicit_vs_implicit <- delta_comparisons %>%
  pull(explicit_vs_implicit)

hist(myVec_explicit_vs_implicit)

aida::summarize_sample_vector(myVec_explicit_vs_implicit)
```
95% credible interval includes 0 -> no evidence for a difference between explicit and implicit content

```{r}
myVec_at_issue_vs_nolonger <- delta_comparisons %>%
  pull(explicit_vs_nolonger)

hist(myVec_at_issue_vs_nolonger)

aida::summarize_sample_vector(myVec_at_issue_vs_nolonger)
```
95% credible interval includes 0 -> no evidence for a difference between at-issue and not-at-issue 


```{r}

myVec_nolonger_vs_never <- delta_comparisons %>%
  pull(nolonger_vs_never)

hist(myVec_nolonger_vs_never)

aida::summarize_sample_vector(myVec_nolonger_vs_never)
```
shows that there is very little difference between no longer at issue content (clausal NAI constructions) and never at issue content (presuppositions etc) -> suggests there is no difference within not-at-issue constructions 


95% credible interval includes 0 -> no evidence for a difference 


```{r}

myVec_at_issue_vs_never <- delta_comparisons %>%
  pull(explicit_vs_never)

hist(myVec_at_issue_vs_never)

aida::summarize_sample_vector(myVec_at_issue_vs_never)
```

CI includes 0 -> no evidence for effect (however this is the closest we got to it being not 0)


SAMPLE SIZE RATIONALE - SIMULATIONS WITH DIFFERENT PARTICIPANT SIZES:

idea: simulate 1000 experiments with 50 participants to see in how many of the 1000 experiments we would find effects.


```{r}
#define participant number to test

N <- 50

items <- unique(df_main$itemID)
```



```{r}
#set up design so that each participant sees all 8 conditions across 8 randomly sampled items

conditions <- expand_grid(
  informationStatus = c(
    "at_issue_explicit",
    "at_issue_implicit",
    "no_longer_at_issue",
    "never_at_issue"
  ),
  contextType = c(
    "positive",
    "negative"
  )
)


set.seed(123)

design <- map_dfr(
  1:N,
  function(p) {

    sampled_items <- sample(
      items,
      size = 8,
      replace = FALSE
    )

    conditions %>%
      mutate(
        itemID = factor(
          sample(sampled_items),
          levels = items
        ),
        submission_id = factor(p)
      )
  }
) %>%
  mutate(
    informationStatus = factor(
      informationStatus,
      levels = levels(df_main$informationStatus)
    ),
    contextType = factor(
      contextType,
      levels = levels(df_main$contextType)
    )
  )

```

```{r}
str(design$informationStatus)
str(design$contextType)
```


```{r}

#generate 1000 posterior predictive draws -> predictions are the simulated observed ratings from 0 - 7 
#using the posterior predictive distribution of the pilot model to generate 50 x 8 new data points

model <- readRDS("pilot_model.rds")

pred <- design %>%
  add_predicted_draws(
    model,
    ndraws = 1,
    re_formula = NULL,
    allow_new_levels = TRUE
  )

print(pred)

```

```{r}
pred %>%
  group_by(submission_id) %>%
  summarise(mean_rating = mean(as.numeric(as.character(.prediction))))
```

```{r}
pred <- pred %>%
  mutate(rating = as.numeric(.prediction))
```


```{r}

#fit a new brms model on the predicted dataset

sim_model <- brm(
  rating ~ informationStatus * contextType +
    (1 + informationStatus * contextType | submission_id) +
    (1 | itemID),
  data = pred,   # -> the simulated dataset
  family = cumulative(link = "logit"),
  chains = 4,
  cores = 4,
  iter = 2000
)
```


```{r}
# create condition grid for posterior predictions

newdat <- tidyr::expand_grid(
  informationStatus = c(
    "at_issue_explicit",
    "at_issue_implicit",
    "no_longer_at_issue",
    "never_at_issue"
  ),
  contextType = c(
    "positive",
    "negative"
  )
)

newdat <- newdat %>%
  mutate(
    informationStatus = factor(
      informationStatus,
      levels = levels(df_main$informationStatus)
    ),
    contextType = factor(
      contextType,
      levels = levels(df_main$contextType)
    )
  )

```


```{r}
#get posterior predictions (epreds)

draws <- newdat %>%
  add_epred_draws(
    sim_model,
    re_formula = NA
  )

```



```{r}
#compute expected rating

draws_expected <- draws %>%
  mutate(
    category_num = as.numeric(as.character(.category))
  ) %>%
  group_by(
    .draw,
    informationStatus,
    contextType
  ) %>%
  summarise(
    expected_rating =
      sum(category_num * .epred),
    .groups = "drop"
  )
```


```{r}
#compute the contrasts (differences of differences)

contrasts <- draws_expected %>%
  mutate(
    condition = paste(
      informationStatus,
      contextType,
      sep = "_"
    )
  ) %>%
  select(
    .draw,
    condition,
    expected_rating
  ) %>%
  pivot_wider(
    names_from = condition,
    values_from = expected_rating
  ) %>%
  mutate(

    atissue_vs_never =
      (
        at_issue_explicit_positive -
        at_issue_explicit_negative
      ) -
      (
        never_at_issue_positive -
        never_at_issue_negative
      ),

    atissue_vs_implicit =
      (
        at_issue_explicit_positive -
        at_issue_explicit_negative
      ) -
      (
        at_issue_implicit_positive -
        at_issue_implicit_negative
      ),

    nolonger_vs_never =
      (
        no_longer_at_issue_positive -
        no_longer_at_issue_negative
      ) -
      (
        never_at_issue_positive -
        never_at_issue_negative
      )
  )
```


```{r}

#define effect function

detect_effect <- function(x) {

  ci <- quantile(
    x,
    probs = c(.025, .975)
  )

  ci[1] > 0 || ci[2] < 0
}


```


```{r}
#test posterior predictions for effect:

tibble(
  atissue_vs_never =
    detect_effect(contrasts$atissue_vs_never),

  atissue_vs_implicit =
    detect_effect(contrasts$atissue_vs_implicit),

  nolonger_vs_never =
    detect_effect(contrasts$nolonger_vs_never)
)
```

```{r}
summarise_contrast <- function(x) {
  tibble(
    mean = mean(x),
    lower = quantile(x, .025),
    upper = quantile(x, .975)
  )
}

summarise_contrast(contrasts$atissue_vs_never)
summarise_contrast(contrasts$atissue_vs_implicit)
summarise_contrast(contrasts$nolonger_vs_never)
```


```{r}


run_simulation <- function(
  pilot_model,
  items,
  N = 100
) {

  # -----------------------------
  # Generate design
  # -----------------------------

  conditions <- tidyr::expand_grid(
    informationStatus = c(
      "at_issue_explicit",
      "at_issue_implicit",
      "no_longer_at_issue",
      "never_at_issue"
    ),
    contextType = c(
      "positive",
      "negative"
    )
  )

  design <- purrr::map_dfr(
    1:N,
    function(p) {

      sampled_items <- sample(
        items,
        size = 8,
        replace = FALSE
      )

      conditions %>%
        mutate(
          submission_id = factor(p),
          itemID = factor(
            sample(sampled_items),
            levels = items
          )
        )
    }
  ) %>%
    mutate(
      informationStatus = factor(
        informationStatus,
        levels = levels(df_main$informationStatus)
      ),
      contextType = factor(
        contextType,
        levels = levels(df_main$contextType)
      )
    )

  # -----------------------------
  # Simulate ratings
  # -----------------------------

  pred <- design %>%
    tidybayes::add_predicted_draws(
      pilot_model,
      ndraws = 1,
      re_formula = NULL,
      allow_new_levels = TRUE
    ) %>%
    mutate(
      rating = as.numeric(.prediction)
    )

  # -----------------------------
  # Fit analysis model
  # -----------------------------

  sim_model <- brm(
    rating ~ informationStatus * contextType +
      (1 + informationStatus * contextType | submission_id) +
      (1 | itemID),
    data = pred,
    family = cumulative(link = "logit"),
    chains = 4,
    cores = 4,
    iter = 2000,
    warmup = 1000,
    refresh = 0
  )

  # -----------------------------
  # Posterior predictions
  # -----------------------------

  newdat <- tidyr::expand_grid(
    informationStatus = levels(df_main$informationStatus),
    contextType = levels(df_main$contextType)
  )

  draws <- newdat %>%
    tidybayes::add_epred_draws(
      sim_model,
      re_formula = NA
    )

  # -----------------------------
  # Expected ratings
  # -----------------------------

  draws_expected <- draws %>%
    mutate(
      category_num =
        as.numeric(as.character(.category))
    ) %>%
    group_by(
      .draw,
      informationStatus,
      contextType
    ) %>%
    summarise(
      expected_rating =
        sum(category_num * .epred),
      .groups = "drop"
    )

  # -----------------------------
  # Contrasts
  # -----------------------------

  contrasts <- draws_expected %>%
    mutate(
      condition = paste(
        informationStatus,
        contextType,
        sep = "_"
      )
    ) %>%
    select(
      .draw,
      condition,
      expected_rating
    ) %>%
    pivot_wider(
      names_from = condition,
      values_from = expected_rating
    ) %>%
    mutate(

      atissue_vs_never =
        (
          at_issue_explicit_positive -
          at_issue_explicit_negative
        ) -
        (
          never_at_issue_positive -
          never_at_issue_negative
        ),

      explicit_vs_implicit =
        (
          at_issue_explicit_positive -
          at_issue_explicit_negative
        ) -
        (
          at_issue_implicit_positive -
          at_issue_implicit_negative
        ),

      nolonger_vs_never =
        (
          no_longer_at_issue_positive -
          no_longer_at_issue_negative
        ) -
        (
          never_at_issue_positive -
          never_at_issue_negative
        )
    )
  
  extract_effect <- function(x) {
  
    hdi <- bayestestR::hdi(
      x,
      ci = 0.95
    )
  
    tibble(
      mean = mean(x),
      median = median(x),
      hdi_low = hdi$CI_low,
      hdi_high = hdi$CI_high,
      detected =
        hdi$CI_low > 0 ||
        hdi$CI_high < 0
    )
  }
  
  
    e1 <- extract_effect(
      contrasts$atissue_vs_never
    )
    
    e2 <- extract_effect(
      contrasts$explicit_vs_implicit
    )
    
    e3 <- extract_effect(
      contrasts$nolonger_vs_never
    )
    
    
    tibble(
    
      atissue_vs_never_detected =
        e1$detected,
      atissue_vs_never_mean =
        e1$mean,
      atissue_vs_never_hdi_low =
        e1$hdi_low,
      atissue_vs_never_hdi_high =
        e1$hdi_high,
    
      explicit_vs_implicit_detected =
        e2$detected,
      explicit_vs_implicit_mean =
        e2$mean,
      explicit_vs_implicit_hdi_low =
        e2$hdi_low,
      explicit_vs_implicit_hdi_high =
        e2$hdi_high,
    
      nolonger_vs_never_detected =
        e3$detected,
      nolonger_vs_never_mean =
        e3$mean,
      nolonger_vs_never_hdi_low =
        e3$hdi_low,
      nolonger_vs_never_hdi_high =
        e3$hdi_high
    )
        
    
}
```



```{r}
model <- readRDS("pilot_model.rds")

set.seed(123)

results <- purrr::map_dfr(
  1:250,
  ~ run_simulation(
      pilot_model = model,
      items = items,
      N = 100
    )
)

```




```{r}
bootstrap_mean_ci <- function(x, R = 5000) {

  x <- x[!is.na(x)]

  boot_means <- replicate(
    R,
    mean(sample(x, replace = TRUE))
  )

  list(
    mean = mean(x),
    lower = as.numeric(quantile(boot_means, 0.025)),
    upper = as.numeric(quantile(boot_means, 0.975))
  )
}

```

```{r}
make_summary_table <- function(results) {

  contrasts <- c(
    "atissue_vs_never",
    "explicit_vs_implicit",
    "nolonger_vs_never"
  )

  purrr::map_dfr(contrasts, function(con) {

    mean_col <- paste0(con, "_mean")
    det_col  <- paste0(con, "_detected")

    boot_ci <- bootstrap_mean_ci(results[[mean_col]])

    tibble::tibble(
      contrast = con,

      detection_rate = mean(results[[det_col]]),

      mean_effect = boot_ci$mean,
      mean_effect_ci_lower = boot_ci$lower,
      mean_effect_ci_upper = boot_ci$upper,

      sd_across_simulations = sd(results[[mean_col]])
    )
  })
}
```


```{r}

#results table 50 participants, 250 simulations

evaluation_table <- make_summary_table(results)

evaluation_table
```


```{r}

evaluation_table <- make_summary_table(results)

evaluation_table
```


```{r}

evaluation_table <- make_summary_table(results)
 
evaluation_table

```



```{r}
#results 50 participants, 250 sims

results %>%
  summarise(
    atissue_vs_never =
      mean(atissue_vs_never),

    explicit_vs_implicit =
      mean(explicit_vs_implicit),

    nolonger_vs_never =
      mean(nolonger_vs_never)
  )
```

result 50 participants, 250 simulations:
at-issue vs never: 0.84
explicit vs implicit: 0.436
nolonger vs. never: 0.28




```{r}
boot_out <- boot::boot(
  data = data.frame(x = results$atissue_vs_never_mean),
  statistic = function(data, indices) {
    mean(data$x[indices])
  },
  R = 2000
)

str(boot_out$t)
```


```{r}
#results 100 participants

results %>%
  summarise(
    atissue_vs_never =
      mean(atissue_vs_never),

    explicit_vs_implicit =
      mean(explicit_vs_implicit),

    nolonger_vs_never =
      mean(nolonger_vs_never)
  )
```

result 100 participants, 250 simulations:
at-issue vs never: 0.916
explicit vs implicit: 0.544
nolonger vs. never: 0.432


