---
title: "pilot_3_data_analysis"
output: html_document
date: "2026-04-20"
---

```{r}
# load packages

library(tidyverse)
library(ggbeeswarm)
library(brms)
library(tidyboot)
library(tidyjson)
library(tidybayes)
library(patchwork)
library(GGally)
library(cowplot)
library(BayesFactor)
library(aida)   # custom helpers: https://github.com/michael-franke/aida-package
library(faintr) # custom helpers: https://michael-franke.github.io/faintr/index.html
library(cspplot)
library(knitr)
library(ggplot2)
library(dplyr)
library(gtools)
library(Hmisc)
library(brms)

##################################################

# these options help Stan run faster
options(mc.cores = parallel::detectCores(),
        brms.backend = "cmdstanr")

# use the CSP-theme for plotting
theme_set(theme_csp())

# global color scheme from CSP
project_colors = cspplot::list_colors() |> pull(hex)

# setting theme colors globally
scale_colour_discrete <- function(...) {
  scale_colour_manual(..., values = project_colors)
}

scale_fill_discrete <- function(...) {
  scale_fill_manual(..., values = project_colors)
}

##################################################

# open data from pilot3 
data <- read.csv("../pilot_3_data/A1-Indicom-pilot-v3-results-collapsed.csv", sep=";") 

# filter out control -> leave only main trials
df_main <- data[data$condition !='control',]

head(df_main)
```

## 

```{r}

#trim stray whitespace
df_main$contextType       <- trimws(df_main$contextType)
df_main$informationStatus <- trimws(df_main$informationStatus)

#ensure correct types
df_main$rating            <- as.numeric(df_main$rating)
df_main$informationStatus <- factor(df_main$informationStatus, levels = c(
  "at-issue-explicit", "at-issue-implicit", "no-longer-at-issue", "never-at-issue"
))
df_main$contextType <- factor(df_main$contextType, levels = c("positive", "negative"))


#get means, median and mode
mode_val <- function(x) {
  x <- x[!is.na(x)]
  as.numeric(names(sort(table(x), decreasing = TRUE))[1])
}
 
summary_table <- df_main %>%
  group_by(informationStatus, contextType) %>%
  summarise(
    n      = n(),
    mean   = round(mean(rating, na.rm = TRUE), 2),
    median = median(rating, na.rm = TRUE),
    mode   = mode_val(rating),
    sd     = round(sd(rating, na.rm = TRUE), 2),
    .groups = "drop"
  ) %>%
  arrange(informationStatus, contextType)
 
print(summary_table, n = Inf)
write.csv(summary_table, "summary_table.csv", row.names = FALSE)
message("Summary table saved to summary_table.csv")





```



```{r }

df_main$informationStatus <- recode(df_main$informationStatus,
  "at-issue-explicit"  = "at-issue\nexplicit",
  "at-issue-implicit"  = "at-issue\nimplicit",
  "no-longer-at-issue" = "no-longer\nat-issue",
  "never-at-issue"     = "never\nat-issue"
)


```
```{r }
# ── Plot 2: x = informationStatus, grouped by contextType ───────────────────
p2 <- ggplot(df_main, aes(x = informationStatus, y = rating, fill = contextType)) +
  stat_summary(
    fun = mean,
    geom = "bar",
    position = position_dodge(0.9)
  ) +
  stat_summary(
    fun.data = mean_cl_boot,
    geom = "errorbar",
    width = 0.2,
    position = position_dodge(0.9)
  ) +
  labs(
    x     = "Information Status",
    y     = "Mean Rating",
    fill  = "Context Type",
    title = "Mean Ratings by Information Status and Context Type\n(95% Bootstrapped Confidence Intervals)"
  ) +
  theme_csp()
 
print(p2)
ggsave("plot2_across_informationStatus.png", plot = p2, width = 10, height = 6, dpi = 150)
```






```{r }


# ── Participant mean ratings (overall) ───────────────────────────────────────
pp_summary <- df_main %>%
  group_by(submission_id) %>%
  summarise(
    mean_rating = mean(rating, na.rm = TRUE),
    lower_ci    = smean.cl.boot(rating)[2],
    upper_ci    = smean.cl.boot(rating)[3],
    n           = n(),
    .groups = "drop"
  )
 
ggplot(pp_summary, aes(x = reorder(submission_id, mean_rating), y = mean_rating)) +
  geom_point() +
  geom_errorbar(aes(ymin = lower_ci, ymax = upper_ci), width = 0.2) +
  coord_flip() +
  theme_csp() +
  labs(
    x     = "Participant",
    y     = "Mean Rating",
    title = "Participant Mean Ratings"
  )
 
ggsave("plot_pp_overall.png", width = 8, height = 6, dpi = 150)

```

```{r }
# ── Participant mean ratings by contextType ───────────────────────────────────
pp_context_summary <- df_main %>%
  group_by(submission_id, contextType) %>%
  summarise(
    mean_rating = mean(rating, na.rm = TRUE),
    lower_ci    = smean.cl.boot(rating)[2],
    upper_ci    = smean.cl.boot(rating)[3],
    .groups = "drop"
  )
 
ggplot(pp_context_summary, aes(x = contextType, y = mean_rating)) +
  geom_point() +
  geom_errorbar(aes(ymin = lower_ci, ymax = upper_ci), width = 0.2) +
  facet_wrap(~ submission_id) +
  theme_minimal() +
  labs(
    x     = "Context Type",
    y     = "Mean Rating",
    title = "Participant Mean Ratings by Context Type"
  )
 
ggsave("plot_pp_by_context.png", width = 12, height = 8, dpi = 150)

```



```{r, message=FALSE, warning=FALSE}


 
df_main$informationStatus <- recode(df_main$informationStatus,
  "at-issue\nexplicit"  = "at_issue_explicit",
  "at-issue\nimplicit"  = "at_issue_implicit",
  "no-longer\nat-issue" = "no_longer_at_issue",
  "never\nat-issue"     = "never_at_issue"
)


# ── Model ────────────────────────────────────────────────────────────────────
# informationStatus is the main predictor
# contextType included as a control variable
# random intercepts for participants and items
 
model <- brm(
  #rating ~ informationStatus + contextType + (1 | submission_id) + (1 | itemID),
    rating ~ informationStatus * contextType + (1 + informationStatus * contextType | submission_id) + (1 | itemID),
  data   = df_main,
  family = cumulative(link = "logit"),
  chains = 4,
  cores  = 4,
  iter   = 4000
)
 
summary(model)
saveRDS(model, "pilot_model.rds")
 
```
```{r}

draws <- df_main  %>% 
  select(informationStatus, contextType) %>% 
  unique() %>% 
  tidybayes::add_linpred_draws( #use add_epred_draws() instead? -> expected ratings instead of log odds
    model,
    ndraws = 2000,#number of draws
    re_formula = NA
  )

```


```{r}
deltas <- draws %>%
  ungroup() %>%
  select(.draw, informationStatus, contextType, .linpred) %>%
  pivot_wider(
    names_from = contextType,
    values_from = .linpred
  ) %>%
  mutate(
    delta = positive - negative #compute the difference between positive and negative per conditions
  )
```

```{r}
myVec_at_issue_explicit <- deltas %>%
  filter(informationStatus == "at_issue_explicit") %>%
  pull(delta)

myVec_at_issue_implicit <- deltas %>%
  filter(informationStatus == "at_issue_implicit") %>%
  pull(delta)

myVec_no_longer_at_issue <- deltas %>%
  filter(informationStatus == "no_longer_at_issue") %>%
  pull(delta)

myVec_never_at_issue <- deltas %>%
  filter(informationStatus == "never_at_issue") %>%
  pull(delta)
```


```{r}
hist(myVec_at_issue_explicit)
hist(myVec_at_issue_implicit)
hist(myVec_no_longer_at_issue)
hist(myVec_never_at_issue)

aida::summarize_sample_vector(myVec_at_issue_explicit)
aida::summarize_sample_vector(myVec_at_issue_implicit)
aida::summarize_sample_vector(myVec_no_longer_at_issue)
aida::summarize_sample_vector(myVec_never_at_issue)


```
```{r}
delta_comparisons <- deltas %>%
  select(.draw, informationStatus, delta) %>%
  pivot_wider(
    names_from = informationStatus,
    values_from = delta
  ) %>%
  mutate(
    explicit_vs_nolonger =
      at_issue_explicit - no_longer_at_issue,

    explicit_vs_implicit =
      at_issue_explicit - at_issue_implicit,

    nolonger_vs_never =
      no_longer_at_issue - never_at_issue,
    
    explicit_vs_never = 
      at_issue_explicit - never_at_issue
  )
```

```{r}
myVec_explicit_vs_implicit <- delta_comparisons %>%
  pull(explicit_vs_implicit)

hist(myVec_explicit_vs_implicit)

aida::summarize_sample_vector(myVec_explicit_vs_implicit)
```
95% credible interval includes 0 -> no evidence for a difference between explicit and implicit content

```{r}
myVec_at_issue_vs_nolonger <- delta_comparisons %>%
  pull(explicit_vs_nolonger)

hist(myVec_at_issue_vs_nolonger)

aida::summarize_sample_vector(myVec_at_issue_vs_nolonger)
```
95% credible interval includes 0 -> no evidence for a difference between at-issue and not-at-issue 


```{r}

myVec_nolonger_vs_never <- delta_comparisons %>%
  pull(nolonger_vs_never)

hist(myVec_nolonger_vs_never)

aida::summarize_sample_vector(myVec_nolonger_vs_never)
```
shows that there is very little difference between no longer at issue content (clausal NAI constructions) and never at issue content (presuppositions etc) -> suggests there is no difference within not-at-issue constructions 


95% credible interval includes 0 -> no evidence for a difference 


```{r}

myVec_at_issue_vs_never <- delta_comparisons %>%
  pull(explicit_vs_never)

hist(myVec_at_issue_vs_never)

aida::summarize_sample_vector(myVec_at_issue_vs_never)
```

CI includes 0 -> no evidence for effect (however this is the closest we got to it being not 0)

----

SAMPLE SIZE RATIONALE - HOW MANY PARTICIPANTS DO WE NEED

idea: 



```{r}
#define participant number to test

N <- 50

items <- unique(df_main$itemID)
```



```{r}
#set up design so that each participant sees all 8 conditions across 8 randomly sampled items

set.seed(123)

design <- map_dfr(
  1:N,
  function(p) {
    
    sample_items <- sample(items, 8, replace = FALSE)
    
    expand_grid(
      submission_id = factor(p),
      itemID = factor(sample_items, levels = items),
      informationStatus = c(
        "at_issue_explicit",
        "at_issue_implicit",
        "no_longer_at_issue",
        "never_at_issue"
      ),
      contextType = c("positive", "negative")
    )
  }
)

design <- design %>%
  mutate(
    informationStatus = factor(
      informationStatus,
      levels = levels(df_main$informationStatus)
    ),
    contextType = factor(
      contextType,
      levels = levels(df_main$contextType)
    )
  )


```

```{r}
str(design$informationStatus)
str(design$contextType)
```


```{r}

#generate 1000 posterior predictive draws -> predictions are the simulated observed ratings from 0 - 7 

model <- readRDS("pilot_model.rds")

pred <- design %>%
  add_predicted_draws(
    model,
    ndraws = 1000,
    re_formula = NULL,
    allow_new_levels = TRUE
  )


```


```{r}
pred <- pred %>%
  mutate(
    .prediction = as.numeric(as.character(.prediction))
  )
```



```{r}
#compute within-participant polarity effects (the deltas)
#positive - negative per participant x condition

deltas <- pred %>%
  select(
    .draw,
    submission_id,
    itemID,
    informationStatus,
    contextType,
    .prediction
  ) %>%
  pivot_wider(
    id_cols = c(
      .draw,
      submission_id,
      itemID,
      informationStatus
    ),
    names_from = contextType,
    values_from = .prediction
  ) %>%
  mutate(
    delta = positive - negative
  )
```


```{r}

#aggregate deltas per condition (within each posterior draw)
#this computes an average for each condition within one simulated experiment -> so that we get one value per condition per simulated experiment

delta_summary <- deltas %>%
  group_by(.draw, informationStatus) %>%
  summarise(delta = mean(delta, na.rm = TRUE), .groups = "drop")

```


```{r}

#reshape for contrasts

delta_wide <- delta_summary %>%
  pivot_wider(
    names_from = informationStatus,
    values_from = delta
  )

```

```{r}
delta_summary %>%
  group_by(informationStatus) %>%
  summarise(
    mean = mean(delta),
    sd = sd(delta)
  )
```



```{r}

#compute the three theoretical contrasts

contrasts <- delta_wide %>%
  mutate(
    atissue_vs_never =
      at_issue_explicit - never_at_issue,

    atissue_vs_implicit =
      at_issue_explicit - at_issue_implicit,

    nolonger_vs_never =
      no_longer_at_issue - never_at_issue
  )

```




```{r}
contrasts %>%
  summarise(across(
    c(atissue_vs_never, atissue_vs_implicit, nolonger_vs_never),
    ~ quantile(.x, c(.025, .5, .975), na.rm = TRUE)
  ))
```



```{r}

#define the criterion -> 95% credible interval excludes 0

detect_effect <- function(x) {
  ci <- quantile(x, c(.025, .975))
  (ci[1] > 0) | (ci[2] < 0)
}

```



```{r}
posterior_summary <- tibble(
  contrast = c(
    "atissue_vs_never",
    "atissue_vs_implicit",
    "nolonger_vs_never"
  ),
  p_gt_0 = c(
    mean(contrasts$atissue_vs_never > 0),
    mean(contrasts$atissue_vs_implicit > 0),
    mean(contrasts$nolonger_vs_never > 0)
  ),
  mean = c(
    mean(contrasts$atissue_vs_never),
    mean(contrasts$atissue_vs_implicit),
    mean(contrasts$nolonger_vs_never)
  ),
  ci_lower = c(
    quantile(contrasts$atissue_vs_never, .025),
    quantile(contrasts$atissue_vs_implicit, .025),
    quantile(contrasts$nolonger_vs_never, .025)
  ),
  ci_upper = c(
    quantile(contrasts$atissue_vs_never, .975),
    quantile(contrasts$atissue_vs_implicit, .975),
    quantile(contrasts$nolonger_vs_never, .975)
  )
)

posterior_summary
```




```{r}
delta_summary %>%
  group_by(informationStatus) %>%
  summarise(
    mean = mean(delta),
    sd = sd(delta),
    q025 = quantile(delta, 0.025),
    q975 = quantile(delta, 0.975)
  )
```


